Variational Monte Carlo on bosonic systems

### 2018-03-02 project_1/article/article.tex

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -2\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
    \end{align}
    It is worth noting that for $\alpha = \pm\half\omega$ we will find a stable
    value which turns out to be the exact energy minimum. This happens as the
    entire sum over all the random walkers disappears.

>>>>

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -2\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the entire sum over all the random walkers
    disappears.

    \subsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}.
        \end{align}

        % TODO: Add the computation of this expression
        The final result for the energy is
        \begin{align}
            E(\alpha)
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
        \end{align}
        By minimizing this expression, i.e., finding the derivative of the
        energy with respect to $\alpha$ and equating this to zero, yields the
        expected minimum of variational energy to be
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}
        which in natural units reduces to $\alpha_0 = \half\omega$. The energy at
        this value of $\alpha$ (in natural units) is then
        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2}.
        \end{align}
